Forecasting Gamma-Ray Bursts using Gravitational Waves

### 2018-04-03 Paper/Manuscript.tex
Finished 2nd read through

\title{Forecasting Gamma-ray Bursts using Gravitational Waves}

@@

\section{Introduction}\label{Sec:Intro}
The 6.95\footnote{There have thus far been six detections with $5\sigma$ statistical significance and one additional event, LVT151012 \cite{TheLIGOScientific:2016pea}, with $2\sigma \simeq 95\%$ significance hence 6.95 overall.} 
gravitational-wave merger events detected by the Advanced Ligo-Virgo network have firmly established gravitational-wave astronomy
as an observational science \cite{TheLIGOScientific:2016pea, Abbott:2017gyy, Abbott:2017oio, Abbott:2017vtc, GW170817}. 
Though the first event of 14 September 2015 (GW150914 \cite{GW150914}) will always be the ``poster-child'' of this field,
the icing on the cake was the 17 August 2017 (GW170817 \cite{GW170817}) event involving the inspiral and merger of a binary neutron star system 
promptly followed, 1.7 seconds later, by a gamma-ray burst (GRB170817A) detected by the Fermi Gamma-ray Burst Monitor \cite{Fermi} and by the International Gamma-Ray Astrophysics Laboratory \cite{Svinkin, Savchenko:2017ffs}.
The Advanced-LIGO-Virgo network's initial source localization to within $\sim 31\,\text{degree}^2$ in the Southern skies
enabled astronomers to locate the electromagnetic counterpart in the galaxy NGC 4993, 
first picked up by the One-Meter, Two Hemisphere team less than 11 hours after the merger \cite{Coulter2017, Coulter:2017wya}.
This subsequently launched the largest-ever campaign of multi-messenger astronomy across the entire electromagnetic (EM) spectrum which is still ongoing
more than half a year after the initial GRB \cite{GBM:2017lvd}.

GW170817 was initially identified by the LIGO-Hanford interferometer (H1) 
using a template bank of gravitational waveforms computed within the framework of post-Newtonian theory \cite{Buonanno:2009zt, Blanchet_LRR}.
The inspiral swept across the Advanced Ligo-Virgo (ALV) network's bandwidth from $\sim 30\,$Hz to $\sim 2\,$kHz in 57 seconds
executing $\sim 3000$ gravitational wave cycles and accumulating a network signal-to-noise ratio of 32.4 with a false-alarm rate of one per $8.0\times 10^4$ years \cite{GW170817}. 
The total mass of the system was inferred to be $2.74^{+0.04}_{-0.01} M_\odot$ with component masses in the range $1.17-1.60 M_\odot$ consistent
with neutron stars (assuming low spin priors, see Sec.~\ref{sec:idealizations}).
Moreover, both the gravitational wave (GW) and EM observations are consistent with a source luminosity distance of $\sim 40\,$Mpc
and support the hypothesis that GW170817 resulted from the inspiral and merger of two neutron stars, 
causing the prompt short-hard gamma-ray burst (GRB170817A) in NGC4993 as first suggested by Ref.~\cite{Eichler:1989ve}.

A week after GW170817, the ALV network was taken offline for upgrades for the third observation run (O3) 
scheduled to start in Autumn 2018 with roughly twice the sensitivity of O2.
In addition, the Japanese cryogenic interferometer KAGRA (\cite{KAGRA, KAGRA2}) will start its test runs in 2018-2019 and reach design sensitivity circa 2021-2022 \cite{Akutsu:2017thy, Aasi:2013wya}. By 2025, LIGO-India detector will join the global interferometer network hence improving both the overall sensitivity and the sky-localization capabilities of the interferometer network \cite{Aasi:2013wya}.
Around this time, the construction of the first third-generation ground-based interferometer, Einstein Telescope, will begin in Europe \cite{ET_doc}. 
Meanwhile, the US is currently considering a mid-2020s update to LIGO called Voyager \cite{LIGO_Voy} and the construction of a more ambitious 40\,km long interferometer dubbed The Cosmic Explorer \cite{CE}.
All of these near-future improvements to the GW-detector network beg a simple question: can we detect the inspiral of a binary neutron star system
early enough to witness a gamma-ray burst as it happens? In other words, can we forecast GRBs using GW detections?

@@

Given the sensitivities of interferometers to date, efforts have thus far mostly focused on reducing the latency of EM follow-ups, 
i.e., the lag time between the merger and the alerting of the EM observatories. 
LIGO-Virgo science runs in 2009-2010 yielded a latency of $\sim 30$ to $60\,$minutes \cite{Abbott:2011ys} (no detections).
In the case of GW170817, the latency was 40\,minutes and $36^{+1.7}\,$seconds \cite{GBM:2017lvd}.
It then took an additional $\sim 10$ hours to locate the optical transient. %designated AT 2017gfo.

Here, we are interested  in exploring the predictive power of near-future ground-based GW detectors. 
%thus leave the task of reducing the latency to the experts.
More specifically, we wish to find out how much early warning the GW detectors can provide us before the merger/GRB.
Ref.~\cite{Cannon:2011vi} approached this from an algorithmic perspective by developing computationally inexpensive methods to detect inspiral
signals in GW data to provide early-warning triggers \emph{before} the merger (also see Ref.~\cite{Knowles:2018hqq} for a very recent improvement to what is currently being
used in the ALV pipeline).
We instead focus on the forecasting capabilities of the ground-based interferometers at their \emph{full design} sensitivity by computing the time difference
between the merger and the instant of detection (detector trigger) defined in terms of a certain threshold signal-to-noise ratio.
To this end, we consider the ALV network of 2020s composed of three L-shaped interferometers operating at their design sensitivities
and Einstein Telescope of 2030s made up of triangular interferometers.
Ref.~\cite{Nissanke:2012dj} investigated the capabilities of three to five-interferometer detector networks of 2020s in terms of their horizon distances
to inspiralling BNSs and their sky-localization of these sources. Our work here is somewhat complementary in the sense that we are mostly concerned with
how early the GW networks can detect the inspiral before the merger.

@@

Our computation for $T_\text{AW}$ is based on the Newtonian evolution of a binary system composed of two point masses in a quasi-circular Keplerian orbit. 
The long-term dynamics is governed by the radiation reaction of 
the GW emission from the system due to the time variance of the quadrupole moment of the binary.
Though the Newtonian (leading-order) evolution is better than $ 95\%$ accurate
we nonetheless supplement it with results from post-Newtonian theory up to 3.5 post-Newtonian order \cite{Blanchet_LRR}.
The evolution of a BNS system in the strong-field regime, where speeds $\gtrsim 0.1c$, is a very complex and complicated problem requiring fully relativistic, magnetohydrodynamic codes 
with neutrino transport running on large-scale computing clusters and taking up to millions of CPU hours 
(see Ref.~\cite{Faber:2012rw} for a review, and Refs.~\cite{Kyutoku:2017voj, Zappa:2017xba, Dietrich:2018upm} for the latest developments and references therein).
Despite its simplicity, our quadrupole radiation driven adiabatic
evolution is more than sufficient to provide reliable estimations of advance warning times.
We support this claim in Sec.~\ref{sec:idealizations} with an extensive list of computations showing how much each of our approximations affects the inspiral time.

@@

This article is organized as follows. Sec.~\ref{sec:BNS_inspiral} introduces the formulation for the leading-order evolution of the binary inspiral.
Sec.~\ref{sec:IFO_response} details the interferometer response to GWs based on topology.
Sec.~\ref{sec:idealizations} lists the various idealizations we employ to simplify our treatment and supplies justification for each.
Sec.~\ref{sec:GW170817} tests our model network and evolution using the parameters of GW170817 and the corresponding GW data.
Sec.~\ref{sec:results} contains our main results presented as advance warning times in Tables~\ref{table:LIGO2020}, \ref{table:ET}; 
and as horizon distances and event rates in Table~\ref{table:horizon}.
In Sec.~\ref{sec:BH_NS}, we recompute the advance warning times in the case of black hole - neutron star inspirals to see whether or not 
future detectors can forecast a tidal disruption event, for which we summarize our findings in Table~\ref{table:ET_BH_NS}.

Throughout the text, $f$ denotes the GW frequency of the dominant quadrupole mode, which is twice the orbital (Keplerian) frequency $f_K$. 
%Relatedly, we define the following angular frequencies: $\omega = 2\pi f, \Omega=2\pi f_K$ with $\omega=2\Omega$.
We employ the $\simeq$ symbol when displaying the numerical values of quantities %with infinite decimal expansions 
which we usually truncate at four significant digits.
The $\approx$ symbol is reserved for approximations whereas $\sim$ denotes rough, order-of-magnitude equalities,
e.g., $\pi \simeq 3.142$, $\pi \approx 3$, $\pi \sim \ord(1)$. %Kepler's third law: $f_K^2\sim r^{-3}$.
Overdots denote time derivative with respect to detector-frame time, e.g., $\dot{E} =dE/dt$, and $\propto$ denotes proportionalities. 
Unless otherwise noted, we use standard SI units.

\section{Binary neutron star inspirals}\label{sec:BNS_inspiral}
\subsection{Newtonian evolution}
For us, the starting point for the evolution of a binary under gravitational radiation reaction
%the emission of GWs 
is the expression for the power emitted in GWs given by 
%for the generated by the time-variance of the mass quadrupole moment
%
\be
\dot{E} = \f{G}{5c^5}\langle \dddot{Q}_{ij}\dddot{Q}_{ij} \rangle \label{eq:Edot_quadrupole}\, .
\ee
%
This is the celebrated Einstein quadrupole formula \cite{1918SPAW}. 
Here, $Q_{ij}$ is a trace-reversed mass quadrupole moment which
is the dominant contribution to the power emitted %in gravitational radiation
by a system with changing mass moments.

Here, we are specifically interested in the motion of two point masses in bound motion around a common center of mass.
Labelling the masses by $m_1 \le m_2$, we have, for a binary in a circular orbit with Keplerian angular frequency $\Omega$ and separation $2r$,
%
\be
\dot{E} = \f{32}{5}\f{G\mu^2}{c^5}\, r^4 \Omega^6\label{eq:Edot},
\ee
%
where $\mu= m_1 m_2/M$ is the reduced mass and $M=m_1+m_2$ is the total mass. 
We justify our simplification of motion to circular orbits in Sec.~\ref{sec:idealizations}.
%Our assumption of the circularity of the motion is well justified in

An important quantity in GW astronomy is the chirp mass of the binary given by
%
\be
M_c = \mu^{3/5} M^{2/5} = \f{(m_1 m_2)^{3/5}}{(m_1+m_2)^{1/5}} \label{eq:chirp_mass}\, ,
\ee
%
which turns Eq.~(\ref{eq:Edot}) into
%
\be
\dot{E} = \f{32}{5}\f{c^5}{G} \left(\f{G M_c\, \omega}{2c^3}\right)^{10/3}, \label{eq:Edot2}
\ee
%
where $\omega=2\Omega$ is the angular frequency of the quadrupolar GWs. 
The quantity in parantheses in Eq.~(\ref{eq:Edot2}) is ubiquitous in post-Newtonian theory and is
referred to as the dimensionless inverse separation $x$.
As $x\sim v^2/c^2$, it is used to keep track of post-Newtonian orders and we can see from in Eq.~(\ref{eq:Edot2}) that $\dot{E}\sim x^5$. 
%hence the common statement in the literature saying that gravitational radiation starts

The emission of power in GWs decreases the total energy of the binary thus making the orbit more bound. 
The total (binding) energy of the binary is given by
%
\be
E_b = -\frac{G m_1 m_2}{2r}\,  \label{eq:binding_energy1}
\ee
%
giving $\dot{E}_b \propto \dot{r}/r$; hence $\dot{E}_b < 0$ implies $\dot{r}<0$, i.e, the orbital separation decreases in time.
Using Kepler's third law, $ \Omega^2 = G M r^{-3}$, and Eq.~(\ref{eq:chirp_mass}) we obtain
%
\be
E_b = -\left(\f{G^2 M_c^5 \omega^2}{32}\right)^{1/3}\, . \label{eq:binding_energy2}
\ee
%
Energy conservation dictates that $\dot{E}=-\dot{E}_b$. Using the chain rule to determine $\dot{E}_b$ from Eq.~(\ref{eq:binding_energy2})
and setting the resulting expression equal to Eq.~(\ref{eq:Edot2}) yields $\dot{\omega}$ as a function of $\omega$ and various constants.
Translating this expression using $\omega=2\pi f$ we arrive at the standard expression for the GW frequency evolution
%%
%
\be
\dot{f} = \f{96}{5}\pi^{8/3} \f{(G M_c)}{c^5}^{5/3}\, f^{11/3} \label{eq:fdot}
\ee
%
This can be integrated straightforwardly after defining a new time variable $\tau\equiv t_\text{coal}-t$ that equals zero when the neutron stars coalesce.
We can now define an inspiral time at a given GW frequency
%
\be
\tau_\text{insp}(f) = \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,f^{-8/3}\label{eq:tau_insp}\, ,
\ee
%
%
%\be
%\tau_\text{insp}(f) \equiv \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-8/3}-f_\text{ISCO}^{-8/3}\right]\label{eq:tau_insp} \, .
%\ee
%
which we can rewrite as follows
%
\be
\tau_\text{insp}(f) \simeq 16.72\,\text{minutes} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,\left(\f{10\,\text{Hz}}{f}\right)^{8/3}, \label{eq:tau_insp2}
\ee
%
where $1.219 M_\odot$ is the chirp mass corresponding to $m_1=m_2=1.4 M_\odot$.

@@

$\Phi_\text{N}(t)$ in Eqs.~(\ref{eq:hplus_TD}, \ref{eq:hcross_TD}) is the phase of the GWs given by
%
\begin{align}
\Phi_\text{N}(t) &=\int_{t_i}^t dt' \omega(t') \nn\\
&= -2\left(\f{5 G M_c}{c^3}\right)^{-5/8} (t_\text{coal}-t)^{5/8}+ \Phi_0 \, ,\label{eq:Phase}
\end{align}
%
where $\Phi_0 \equiv \Phi(t=t_\text{coal})$ is an integration constant. The subscript N denotes the Newtonian (leading-order) contribution. 
Higher-order contributions can be added in terms a post-Newtonian series, which make up $\lesssim 2\%$ of the total phase.
Although we will include contributions up to and including 3.5PN to obtain the results of Sec.~\ref{sec:results}, 
we will not show these expressions here, but they can be found in Ref.~\cite{Blanchet_LRR}.

The nomenclature ``plus'' ($+$) and ``cross'' ($\times$) follows from the effects that passing GWs have on the plane transverse to their direction of propagation.
If we were to conceive of an ``toy'' detector made up of a circular configuration of test-masses lying in the $x$-$y$ plane, a GW propagating along the $z$ direction would stretch/compress the circular arrangement in a $+$ and $\times$ pattern which is the manifestation of the tidal strain of the GWs on the tranverse plane.
%which is why $h_{+},h_{\times}$ are often called GW strains. 
Next, let us briefly explore how an actual detector, i.e., an interferometer, responds to passing GWs.

@@

To compute the advance warning times we need to know how ``loud'' the binaries become as they sweep across the IFO's detection bandwidth.
This is given by the signal-to-noise ratio (SNR) of the GWs in the detectors' data stream.
The SNR is computed in the frequency domain so we Fourier-transform the time-domain strains of Eqs.~(\ref{eq:hplus_TD}) and (\ref{eq:hcross_TD}) to obtain
%
\begin{align}
 \tilde{h}_+(f) &= A \, \f{c}{D}\, \left(\f{G M_c}{c^3}\right)^{5/6}\, e^{i \Psi_+(f)}\, \f{1}{f^{7/6}}\,  \f{1+\cos^2\iota}{2}  ,\label{eq:hplus_FD}\\
 \tilde{h}_\times(f) & = A \, \f{c}{D}\, \left(\f{G M_c}{c^3}\right)^{5/6}\, e^{i \Psi_\times(f)}\, \f{1}{f^{7/6}}\, \cos\iota \label{eq:hcross_FD},
\end{align}
%
%
where $A= \pi^{-2/3} \sqrt{5/24}$.
Explicit expressions for $\Psi_{+,\times}$ show that they are out of phase by $\pi/2$
(the details of the Fourier transformation can be found in Sec.~4.5 of Ref.~\cite{Maggiore}).
Accordingly, the Fourier transform of the detector strain (\ref{eq:detector_strain}) is given by
%%
%
\be
\tilde{h}(f) = A\, \f{c}{D}\left(\f{G M_c}{c^3}\right)^{5/6} f^{-7/6}\, e^{i\Psi_\text{N}}\, Q(\theta,\phi,\psi,\iota), \label{eq:strain_FD}
\ee
%
%%
where 
%
\be
Q(\theta,\phi,\psi,\iota) = F_+(\theta,\phi,\psi)\f{1+\cos^2\iota}{2}  + i F_\times(\theta,\phi,\psi) \cos\iota \label{eq:Q},
\ee
%
is the quality factor.
The $i$ in front of $F_\times$ in $Q(\theta,\phi,\psi,\iota)$ is due to the $\pi/2$ phase difference between $+$ and $\times$ modes.
$\Psi_\text{N}$ is the leading-order contribution to the frequency-domain phase
%
\be
\Psi_\text{N}=\Psi_+(f)|_\text{N} = 2\pi f (t_\text{coal}+D/c)-\Phi_0-\f{\pi}{4}+ \f{3}{128}\left(\pi f\,\f{G M_c}{c^3}\right)^{-5/3} \label{eq:phase_FD}.
\ee
%
%
Higher-order contributions up to and including 3.5PN can be found in Ref.~\cite{SchutzLRR}.

>>>>

\title{Forecasting Gamma-Ray Bursts using Gravitational Waves}

@@

\section{Introduction}\label{Sec:Intro}
The 6.95\footnote{There have thus far been six detections with $5\sigma$ statistical significance and one additional event, LVT151012 \cite{TheLIGOScientific:2016pea}, with $2\sigma \simeq 95\%$ significance hence 6.95 overall.} 
gravitational-wave merger events detected by the Advanced Ligo-Virgo network have firmly established gravitational-wave astronomy
as an observational science \cite{TheLIGOScientific:2016pea, Abbott:2017gyy, Abbott:2017oio, Abbott:2017vtc, GW170817}. 
Though the first event of 14 September 2015 (GW150914 \cite{GW150914}) will always be the ``poster-child'' of this field,
the icing on the cake was the 17 August 2017 (GW170817 \cite{GW170817}) event involving the inspiral and merger of a binary neutron star system 
promptly followed, 1.7 seconds later, by a gamma-ray burst (GRB170817A) detected by the Fermi Gamma-ray Burst Monitor \cite{Fermi} and by the International Gamma-Ray Astrophysics Laboratory \cite{Svinkin, Savchenko:2017ffs}.
The Advanced-LIGO-Virgo network's initial source localization to within $ 31\,\text{degree}^2$ in the Southern skies
enabled astronomers to locate the electromagnetic counterpart in the galaxy NGC 4993, 
first picked up by the One-Meter, Two Hemisphere team less than 11 hours after the merger \cite{Coulter2017, Coulter:2017wya}.
This subsequently launched the largest-ever campaign of multi-messenger astronomy across the entire electromagnetic (EM) spectrum which is still ongoing
more than half a year after the initial GRB \cite{GBM:2017lvd}.

GW170817 was initially identified by the LIGO-Hanford interferometer (H1) 
using a template bank of gravitational waveforms computed within the framework of post-Newtonian theory \cite{Buonanno:2009zt, Blanchet_LRR}.
The inspiral swept across the Advanced Ligo-Virgo (ALV) network's bandwidth from approximately $ 30\,$Hz to about $ 2\,$kHz in 57 seconds
executing nearly $ 3000$ gravitational wave cycles and accumulating a network signal-to-noise ratio of 32.4 with a false-alarm rate of one per $8.0\times 10^4$ years \cite{GW170817}. 
The total mass of the system was inferred to be $2.74^{+0.04}_{-0.01} M_\odot$ with component masses in the range $1.17-1.60 M_\odot$ consistent
with neutron stars (assuming low spin priors, see Sec.~\ref{sec:idealizations}).
Moreover, both the gravitational wave (GW) and EM observations were consistent with a source luminosity distance of $\sim 40\,$Mpc
and supported the hypothesis that GW170817 resulted from the inspiral and merger of two neutron stars, 
causing the prompt short-hard gamma-ray burst (GRB170817A) in NGC4993 as first suggested by Ref.~\cite{Eichler:1989ve}.

A week after GW170817, the ALV network was taken offline for upgrades for the third observation run (O3) 
scheduled to start in Autumn 2018 with roughly twice the sensitivity of O2.
In addition, the Japanese cryogenic interferometer KAGRA (\cite{KAGRA, KAGRA2}) will start its test runs in 2018-2019 and reach its design sensitivity circa 2021-2022 \cite{Akutsu:2017thy, Aasi:2013wya}. By 2025, LIGO-India detector will join the global interferometer network hence improving both the overall sensitivity and the sky-localization capabilities \cite{Aasi:2013wya}.
Around this time, the construction of the first third-generation ground-based interferometer, Einstein Telescope, will begin in Europe \cite{ET_doc}. 
Meanwhile, the US is currently considering a mid-2020s update to LIGO called Voyager \cite{LIGO_Voy} and the construction of a more ambitious 40\,km long interferometer dubbed The Cosmic Explorer \cite{CE}.
All of these near-future additions to the GW-detector network beg a simple question: can we detect the inspiral of a binary neutron star system
early enough to witness a gamma-ray burst as it happens? In other words, can we forecast GRBs using GW detections?

@@

Given the sensitivities of interferometers to date, efforts have thus far mostly focused on reducing the latency of EM follow-ups, 
i.e., the lag time between the merger and the alerting of the EM observatories. 
LIGO-Virgo science runs in 2009-2010 yielded a latency of $\sim 30$ to $60\,$minutes \cite{Abbott:2011ys} (no detections).
In the case of GW170817, the latency was 40\,minutes and $36^{+1.7}\,$seconds \cite{GBM:2017lvd}.
It then took an additional $ 10$ hours to locate the optical transient. %designated AT 2017gfo.

Here, we are interested  in exploring the predictive power of near-future ground-based GW detectors. 
%thus leave the task of reducing the latency to the experts.
More specifically, we wish to find out how much early warning the GW detectors can provide us before the merger/GRB.
Ref.~\cite{Cannon:2011vi} approached this from an algorithmic perspective by developing computationally inexpensive methods to detect inspiral
signals in GW data to provide early-warning triggers \emph{before} the merger (also see Ref.~\cite{Knowles:2018hqq} for a very recent improvement to what is currently being
used in the ALV pipeline).
We instead focus on the forecasting capabilities of the ground-based interferometers at their \emph{full design} sensitivity by computing the time interval
between the merger and the instant of detection defined in terms of a certain threshold signal-to-noise ratio.
To this end, we consider the ALV network of 2020s composed of three L-shaped interferometers operating at their design sensitivities
and Einstein Telescope of 2030s made up of V-shaped interferometers.
Ref.~\cite{Nissanke:2012dj} investigated the capabilities of three to five-interferometer detector networks of 2020s in terms of their horizon distances
to inspiralling BNSs and their sky-localization of these sources. Our work here is somewhat complementary in the sense that we are mostly concerned with
how early the GW networks can detect the inspiral before the merger.

@@

Our computation for $T_\text{AW}$ is based on the Newtonian evolution of a binary system composed of two point masses in a quasi-circular Keplerian orbit. 
The long-term dynamics is governed by the radiation reaction of 
the GW emission from the system due to the time variance of the quadrupole moment of the binary.
Though the Newtonian (leading-order) treatment gives the 
inspiral time with better than $ 95\%$ accuracy,
we nonetheless supplement it with post-Newtonian corrections up to 3.5 post-Newtonian order (3.5PN) \cite{Blanchet_LRR}.
However, even that fails to provide an accurate depiction of the binary
evolution in the strong-field regime, where speeds $\gtrsim 0.1c$. This is a very complex and complicated problem requiring fully
numerical treatment formulated in terms of general relativistic, magnetohydrodynamic codes 
with neutrino transport running on large-scale computing clusters and taking up to millions of CPU hours 
(see Ref.~\cite{Faber:2012rw} for a review, and Refs.~\cite{Kyutoku:2017voj, Zappa:2017xba, Dietrich:2018upm} for the latest developments and references therein).
Despite its failures, our quadrupole radiation driven Newtonian
evolution is more than sufficient to provide reliable estimations of advance warning times.
We support this claim in Sec.~\ref{sec:idealizations} with an extensive list of computations showing how much each of our approximations affects the inspiral time.

@@

This article is organized as follows. Sec.~\ref{sec:BNS_inspiral} introduces the formulation for the leading-order evolution of the binary inspiral.
Sec.~\ref{sec:IFO_response} details the interferometer response to GWs based on topology.
Sec.~\ref{sec:idealizations} lists the various idealizations we employ to simplify our treatment and supplies justification for each.
Sec.~\ref{sec:GW170817} tests our model network and evolution using the parameters of GW170817 and the corresponding observations.
Sec.~\ref{sec:results} contains our main results presented as advance warning times in Tables~\ref{table:LIGO2020}, \ref{table:ET}; 
and as horizon distances and event rates in Table~\ref{table:horizon}.
In Sec.~\ref{sec:BH_NS}, we recompute the advance warning times in the case of black hole - neutron star inspirals to see whether or not 
future detectors can forecast a tidal disruption event, for which we summarize our findings in Table~\ref{table:ET_BH_NS}.

Throughout the text, $f$ denotes the GW frequency of the dominant quadrupole mode, which is twice the orbital (Keplerian) frequency $f_K$. 
%Relatedly, we define the following angular frequencies: $\omega = 2\pi f, \Omega=2\pi f_K$ with $\omega=2\Omega$.
We employ the $\simeq$ symbol when displaying the numerical values of quantities %with infinite decimal expansions 
which we usually truncate at four significant digits.
The $\approx$ symbol is reserved for approximations whereas $\sim$ denotes rough, order-of-magnitude equalities,
e.g., $\pi \simeq 3.142$, $\pi \approx 3$, $\pi \sim \ord(1)$. %Kepler's third law: $f_K^2\sim r^{-3}$.
Overdots denote time derivatives with respect to detector-frame time, e.g., $\dot{E} =dE/dt$, and $\propto$ denotes proportionalities. 
Unless otherwise noted, we use standard SI units.

\section{Binary neutron star inspirals}\label{sec:BNS_inspiral}
\subsection{Newtonian evolution}
For us, the starting point for the evolution of a binary under gravitational radiation reaction
%the emission of GWs 
is the expression for the power emitted in GWs given by 
the celebrated Einstein quadrupole formula \cite{1918SPAW}
%for the generated by the time-variance of the mass quadrupole moment
%
\be
\dot{E} = \f{G}{5c^5}\langle \dddot{Q}_{ij}\dddot{Q}_{ij} \rangle \label{eq:Edot_quadrupole}\, .
\ee
%
Here, $Q_{ij}$ is a trace-reversed mass quadrupole moment which
is the dominant contribution to the power emitted %in gravitational radiation
by a system with changing mass moments.

Here, we are specifically interested in the motion of two point masses in a circular orbit around a common center of mass.
Labelling the masses by $m_1 \le m_2$, we have, for a binary in a circular orbit with Keplerian angular frequency $\Omega$ and separation $r$,
%
\be
\dot{E} = \f{32}{5}\f{G\mu^2}{c^5}\, r^4 \Omega^6\label{eq:Edot},
\ee
%
where $\mu= m_1 m_2/M$ is the reduced mass and $M=m_1+m_2$ is the total mass. 
We justify our simplification of motion to circular orbits in Sec.~\ref{sec:idealizations}.
%Our assumption of the circularity of the motion is well justified in

An important quantity in GW astronomy is the chirp mass of the binary given by
%
\be
M_c = \mu^{3/5} M^{2/5} = \f{(m_1 m_2)^{3/5}}{(m_1+m_2)^{1/5}} \label{eq:chirp_mass}\, ,
\ee
%
which turns Eq.~(\ref{eq:Edot}) into
%
\be
\dot{E} = \f{32}{5}\f{c^5}{G} \left(\f{G M_c\, \omega}{2c^3}\right)^{10/3}, \label{eq:Edot2}
\ee
%
where $\omega=2\Omega$ is the angular frequency of the quadrupolar GWs. 
The quantity in parantheses in Eq.~(\ref{eq:Edot2}) is ubiquitous in post-Newtonian theory and is
referred to as the dimensionless inverse separation $x$.
As $x\propto v^2/c^2$, it is used to keep track of post-Newtonian orders and we can see from in Eq.~(\ref{eq:Edot2}) that $\dot{E}\propto x^5$. 
%hence the common statement in the literature saying that gravitational radiation starts

The emission of power in GWs decreases the total energy of the binary given by
%
\be
E_b = -\frac{G m_1 m_2}{2r}\,.  \label{eq:binding_energy1}
\ee
%
$\dot{E}_b \propto \dot{r}/r^2$, so $\dot{E}_b < 0$ implies $\dot{r}<0$, i.e, the orbital separation decreases
making the orbit more bound in time.
Using Kepler's third law, $ \Omega^2 = G M r^{-3}$, and Eq.~(\ref{eq:chirp_mass}) we obtain
%
\be
E_b = -\left(\f{G^2 M_c^5 \omega^2}{32}\right)^{1/3}\, . \label{eq:binding_energy2}
\ee
%
Energy conservation dictates that $\dot{E}=-\dot{E}_b$. Using the chain rule to determine $\dot{E}_b$ from Eq.~(\ref{eq:binding_energy2})
and setting the resulting expression equal to Eq.~(\ref{eq:Edot2}) yields $\dot{\omega}$ as a function of $\omega$ and various constants.
Translating this expression using $\omega=2\pi f$ we arrive at the standard expression for the GW frequency evolution
%%
%
\be
\dot{f} = \f{96}{5}\pi^{8/3} \f{(G M_c)}{c^5}^{5/3}\, f^{11/3} \label{eq:fdot}
\ee
%
This can be integrated straightforwardly after defining a new time variable $\tau\equiv t_\text{coal}-t$ that equals zero when the neutron stars coalesce.
Using $\tau$ we can now define an inspiral time at a given GW frequency
%
\be
\tau_\text{insp}(f) = \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,f^{-8/3}\label{eq:tau_insp}\, ,
\ee
%
%
%\be
%\tau_\text{insp}(f) \equiv \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-8/3}-f_\text{ISCO}^{-8/3}\right]\label{eq:tau_insp} \, .
%\ee
%
which we can rewrite as follows
%
\be
\tau_\text{insp}(f) \simeq 16.72\,\text{minutes} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,\left(\f{10\,\text{Hz}}{f}\right)^{8/3}, \label{eq:tau_insp2}
\ee
%
where $1.219 M_\odot$ is the chirp mass corresponding to $m_1=m_2=1.4 M_\odot$.

@@

$\Phi_\text{N}(t)$ in Eqs.~(\ref{eq:hplus_TD}, \ref{eq:hcross_TD}) is the phase of the GWs given by
%
\begin{align}
\Phi_\text{N}(t) &=\int_{t_i}^t dt' \omega(t') \nn\\
&= -2\left(\f{5 G M_c}{c^3}\right)^{-5/8} (t_\text{coal}-t)^{5/8}+ \Phi_0 \, ,\label{eq:Phase}
\end{align}
%
where $\Phi_0 \equiv \Phi(t=t_\text{coal})$ is an integration constant. The subscript N denotes the Newtonian (leading-order) contribution. 
Higher-order contributions can be added in terms a post-Newtonian series, which make up to $\lesssim 2\%$ of the total phase.
Although we will include contributions up to and including 3.5PN to obtain the results of Sec.~\ref{sec:results}, 
we will not show these expressions here, but they can be found in Ref.~\cite{Blanchet_LRR}.

The nomenclature ``plus'' ($+$) and ``cross'' ($\times$) follows from the effects that passing GWs have on the plane transverse to their direction of propagation.
If we were to conceive of a ``toy'' detector made up of a circular configuration of test-masses lying in the $x$-$y$ plane, a GW propagating along the $z$ direction would stretch/compress the circular arrangement in a $+$ and $\times$ pattern 
with a $90^\circ$ phase difference.
This is the manifestation of the tidal strain of the GWs on the tranverse plane.
%which is why $h_{+},h_{\times}$ are often called GW strains. 
Next, let us briefly explore how an actual detector, i.e., an interferometer, responds to passing GWs.

@@

To compute the advance warning times we need to know how ``loud'' the binaries become as they sweep across the IFO's detection bandwidth.
This is given by the signal-to-noise ratio (SNR) of the GWs in the detectors' bandwidth.
The SNR is computed in the frequency domain so we Fourier-transform the time-domain strains of Eqs.~(\ref{eq:hplus_TD}) and (\ref{eq:hcross_TD}) to obtain
%
\begin{align}
 \tilde{h}_+(f) &= A \, \f{c}{D}\, \left(\f{G M_c}{c^3}\right)^{5/6}\, e^{i \Psi_+(f)}\, \f{1}{f^{7/6}}\,  \f{1+\cos^2\iota}{2}  ,\label{eq:hplus_FD}\\
 \tilde{h}_\times(f) & = A \, \f{c}{D}\, \left(\f{G M_c}{c^3}\right)^{5/6}\, e^{i \Psi_\times(f)}\, \f{1}{f^{7/6}}\, \cos\iota \label{eq:hcross_FD},
\end{align}
%
%
where $A= \pi^{-2/3} \sqrt{5/24}$.
Explicit expressions for $\Psi_{+,\times}$ show that they are out of phase by $\pi/2$
(the details of the Fourier transformation can be found in Sec.~4.5 of Ref.~\cite{Maggiore}).
Accordingly, the Fourier transform of the detector strain (\ref{eq:detector_strain}) is given by
%%
%
\be
\tilde{h}(f) = A\, \f{c}{D}\left(\f{G M_c}{c^3}\right)^{5/6} f^{-7/6}\, e^{i\Psi_\text{N}}\, Q(\theta,\phi,\psi,\iota), \label{eq:strain_FD}
\ee
%
%%
where 
%
\be
Q(\theta,\phi,\psi,\iota) = F_+(\theta,\phi,\psi)\f{1+\cos^2\iota}{2}  + i F_\times(\theta,\phi,\psi) \cos\iota \label{eq:Q},
\ee
%
is the quality factor.
The $i$ in front of $F_\times$ in $Q(\theta,\phi,\psi,\iota)$ is due to the $\pi/2$ phase difference between $+$ and $\times$ modes.
$\Psi_\text{N}$ is the leading-order contribution to the frequency-domain phase
%
\be
\Psi_\text{N}=\Psi_+(f)|_\text{N} = 2\pi f (t_\text{coal}+D/c)-\Phi_0-\f{\pi}{4}+ \f{3}{128}\left(\pi f\,\f{G M_c}{c^3}\right)^{-5/3} \label{eq:phase_FD}.
\ee
%
%
Higher-order contributions up to and including 3.5PN can be found in Ref.~\cite{SchutzLRR}.

The detection criterion we employ here is quantified in terms of the optimal SNR defined by
%%
%
\be
\rho=\left[ \int_0^\infty d\ln f\, \f{|2\tilde{h}(f)\sqrt{f}|^2}{S_n(f)}\right]^{1/2} \label{eq:SNR},
\ee
%
%%
where $\sqrt{S_n(f)}$ is the {\it amplitude spectral density} (ASD) of the detector (also called spectral strain sensitivity, or simply, detector noise).
It is usually this quantity that is shown in a typical interferometer strain sensitivity plot with the characteristic strain, $|2\tilde{h}(f)\sqrt{f}|$, 
due to various GW sources overlaid for comparison. 
%
%against the amplitudes of various GW sources whose
%ASDs are given by $2\tilde{h}(f)\sqrt{f}$. This particular combination is called the characteristic strain. 
We show such plots in Secs.~\ref{sec:GW170817} and \ref{sec:results}.
